---
title: "Analyzing and Enhancing Outbreak Response in Ontario Healthcare Institutions: A Comprehensive Study Under the Health Protection and Promotion Act"
author: "Monica Sainani"
thanks: "Code and data are available at: https://github.com/sainanim/Disease-Outbreaks-in-Toronto.git"
date: today
date-format: long
abstract: "This project investigates outbreaks of gastroenteric and respiratory illnesses in Ontario healthcare institutions, focusing on response patterns mandated by the Health Protection and Promotion Act. Through detailed analysis, we aim to uncover insights into outbreak occurrence, durations, and causative agents. Findings will inform strategies to optimize outbreak management and enhance the efficiency of Toronto Public Health's response, contributing to improved public health outcomes."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
---

## 

```{r}
#message: false
#echo: false

#### Preamble ####
# Purpose: Read daata from dataset of Disease Outbreaks in toronto
# Author: Monica Sainani
# Date: 23 January 2024
# Contact: monica.sainani@mail.utoronto.ca
# License: MIT
# Pre-requisites: run scripts/01-download_data.R followed by scripts/02-data_cleaning.R

#### Workspace setup ####
# install.packages("tidyverse")
# install.packages("knitr")
# install.packages("here")
# install.packages("ggplot2")

library(ggplot2)
library(tidyverse)
library(knitr)
library(here)

# read in the data
Disease_data <- read_csv(file = ("outputs/data/Disease_data.csv"),
  show_col_types = FALSE
) 
```

# Introduction

Under the Ontario Health Protection and Promotion Act (HPPA), healthcare institutions are obligated to monitor and report outbreaks of gastroenteric and respiratory illnesses. Toronto Public Health responds to these outbreaks, working collaboratively with healthcare facilities to implement infection prevention and control measures. This project aims to analyze and enhance the response to outbreaks in Ontario healthcare institutions, utilizing a dataset that captures key information under the HPPA.

# Data

## About the dataset

The dataset comprises records of outbreaks in healthcare institutions, including hospitals, long-term care homes, and retirement homes. It includes essential details such as the outbreak setting, types, causative agents, and key dates like the start and end of outbreaks. The data is a valuable resource for understanding the dynamics of infectious outbreaks within healthcare facilities and evaluating the effectiveness of response protocols.

## Preprocessing and data cleaning

The data underwent preprocessing to ensure its suitability for analysis. Steps included handling missing values, converting date columns to the appropriate format, and categorizing variables. Imputation was performed judiciously, and categorical variables were transformed into factors for effective analysis. This rigorous preprocessing lays the foundation for meaningful insights from the dataset.

## Variables of interest

The analysis will focus on key variables, including outbreak settings, types, causative agents, and temporal aspects such as the duration of outbreaks. By investigating these variables, we aim to identify patterns, trends, and factors influencing the occurrence and resolution of outbreaks. The project will contribute to a deeper understanding of outbreak dynamics and guide improvements in response strategies for better public health outcomes.\
\

# Results

```{r}

# Summary statistics
summary(disease_data)

# Exploratory Data Analysis (EDA)
# 1. Outbreak Frequency Over Time
ggplot(disease_data, aes(x = Date.Outbreak.Began)) +
  geom_bar(stat = "count", fill = "skyblue") +
  labs(title = "Outbreak Frequency Over Time", x = "Date", y = "Number of Outbreaks")

##The time series plot reveals patterns in the frequency of outbreaks over time. Peaks or clusters might suggest periods of increased vulnerability or specific events triggering outbreaks, providing valuable information for targeted intervention and resource allocation.

# 2. Outbreak Types Distribution
ggplot(disease_data, aes(x = Type.of.Outbreak, fill = Type.of.Outbreak)) +
  geom_bar() +
  labs(title = "Distribution of Outbreak Types", x = "Outbreak Type", y = "Frequency") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

##The bar plot illustrates the distribution of outbreak types, offering insights into the prevalence of respiratory or gastroenteric outbreaks. This information is crucial for tailoring public health strategies, as different outbreak types may require distinct preventive measures and control strategies.

# 3. Causative Agents Analysis
ggplot(disease_data, aes(x = Causative.Agent.1, fill = Causative.Agent.1)) +
  geom_bar() +
  labs(title = "Distribution of Causative Agents", x = "Causative Agent", y = "Frequency") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

##The bar plot provides an overview of the distribution of causative agents for outbreaks. Identifying the predominant agents can inform targeted surveillance and preventive measures. Knowledge of the primary causative agents is crucial for designing effective public health responses.


```
